---
title: "Cyclistic Bike-Share Analysis Notebook"
author: "Shrabana Nandi"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: false
    highlight: tango
    code_folding: hide
    df_print: paged
    
  pdf_document:
    latex_engine: xelatex
    number_sections: false
    toc: true
    toc_depth: 3
    fig_caption: true
    fig_width: 6
    fig_height: 4
    highlight: tango
    includes:
      in_header: preamble.tex
---


# BUSINESS TASK

**Goal:** Convert casual riders into annual members  

**Core Questions:**
1. How do annual members and casual riders use Cyclistic bikes differently?  
2. Why might casual riders buy annual memberships?  
3. How can Cyclistic use digital media to influence conversion?

🧭 *The plan:* Explore patterns in usage (time, distance, frequency, and stations) and identify behavioral cues that could drive marketing actions.

---

# DATA SOURCE
Cyclistic bike share trip data from Q1 2019 and Q1 2020.

---

# DATA UPLOAD & CLEANING

First, load the essentials. These will handle wrangling, cleaning, and plotting.

```{r setup}
library(tidyverse)
library(lubridate)
library(janitor)
library(skimr)
library(readr)
library(dplyr)
```

💭 *Thoughts:*
No need to go overboard — **tidyverse** covers most cleaning and analysis.
`janitor` will clean column names, and `skimr` gives a quick glance at structure.

---

### Importing Datasets

```{r import}
divvy_2019 <- read_csv("D:/COURSERA/GOOGLE ANALYTICS/COURSE 8/Divvy_Trips_2019_Q1 - Divvy_Trips_2019_Q1.csv")
divvy_2020 <- read_csv("D:/COURSERA/GOOGLE ANALYTICS/COURSE 8/Divvy_Trips_2020_Q1 - Divvy_Trips_2020_Q1.csv")

head(divvy_2019)
head(divvy_2020)
```

💭 *Quick check:*
Just verifying both files load properly — better to catch encoding or column mismatches early.

---

### Cleaning & Standardizing Columns

```{r clean}
divvy_2019 <- clean_names(divvy_2019)
divvy_2020 <- clean_names(divvy_2020)

divvy_2019 <- divvy_2019 %>%
  rename(
    ride_id = trip_id,
    started_at = start_time,
    ended_at = end_time,
    start_station_id = from_station_id,
    start_station_name = from_station_name,
    end_station_id = to_station_id,
    end_station_name = to_station_name,
    member_casual = usertype
  ) %>%
  mutate(ride_id = as.character(ride_id))
```

💭 *Rationale:*
2019 uses older column names, so aligning them with 2020 ensures consistency before merging.
Schema mismatches are common in multi-year data and can break joins if left uncorrected.

---

### Combining Datasets

```{r merge}
common_cols <- intersect(colnames(divvy_2019), colnames(divvy_2020))
divvy_data <- bind_rows(
  divvy_2019 %>% select(all_of(common_cols)),
  divvy_2020 %>% select(all_of(common_cols))
)

rm(divvy_2019, divvy_2020)
skim_without_charts(divvy_data)
```

💭 *Note:*
Merged successfully. Removing old objects frees up memory.
`skim_without_charts()` gives a quick data-health summary — missing values, column types, etc.

---

### Cleaning Nulls, Timestamps, and Duration

```{r cleaning}
divvy_data <- divvy_data %>%
  drop_na() %>%
  mutate(
    started_at = as.POSIXct(started_at, format = "%Y-%m-%d %H:%M:%S"),
    ended_at   = as.POSIXct(ended_at,   format = "%Y-%m-%d %H:%M:%S"),
    ride_length = as.numeric(difftime(ended_at, started_at, units = "mins"))
  ) %>%
  filter(ride_length > 0)
```

💭 *Thinking:*
Dropped nulls because they’re minimal — incomplete records add noise.
Negative or zero durations represent system test rides or logging errors.

---

### Removing Test Rides & Standardizing Labels

```{r filtering}
divvy_data <- divvy_data %>%
  filter(start_station_name != "HQ QR", ride_length < 1440) %>%
  mutate(member_casual = recode(member_casual,
                                "Subscriber" = "member",
                                "Customer"   = "casual"))
```

💭 *Observation:*
“HQ QR” denotes internal QA rides.
A 1-day (1440 min) cutoff removes anomalies.
Renaming “Subscriber/Customer” keeps rider-type labels consistent.

---

### Ride Length Summary

```{r summary}
summary(divvy_data$ride_length)
```

💭 *Interpretation:*
Most rides last **5–15 minutes** (median ≈ 9 min) — short, practical commuting behavior.

---

# DATA ANALYSIS & INSIGHTS

---

## Feature Engineering

```{r features}
divvy_data <- divvy_data %>%
  mutate(
    month = month(started_at, label = TRUE, abbr = TRUE),
    day_of_week = wday(started_at, label = TRUE, abbr = TRUE),
    hour = hour(started_at),
    season = case_when(
      month(started_at) %in% c(12,1,2) ~ "Winter",
      month(started_at) %in% c(3,4,5)  ~ "Spring",
      month(started_at) %in% c(6,7,8)  ~ "Summer",
      TRUE                            ~ "Fall"
    ),
    week_type = ifelse(day_of_week %in% c("Sat","Sun"), "Weekend", "Weekday"),
    ride_category = case_when(
      ride_length <= 5  ~ "Short (<5 min)",
      ride_length <= 15 ~ "Medium (5–15 min)",
      ride_length <= 30 ~ "Long (15–30 min)",
      TRUE              ~ "Extended (>30 min)"
    )
  )
```

💭 *Reasoning:*
Creating temporal and categorical features early simplifies later exploration.
Season, weekday, and duration categories expose behavioral patterns at a glance.

---

## Ride Duration Patterns

```{r duration}
divvy_data %>%
  group_by(member_casual) %>%
  summarise(
    avg_ride_length = mean(ride_length),
    median_ride_length = median(ride_length)
  )
```

💭 *Result:*
Casual riders’ average rides are notably longer — clear leisure orientation.

```{r boxplot-log-keep-minutes, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(scales)

divvy_data %>%
  filter(ride_length > 0) %>%            # remove zeros/negatives
  ggplot(aes(x = member_casual, y = ride_length, fill = member_casual)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +  # hide extreme outliers for visibility
  #geom_jitter(width = 0.15, alpha = 0.12, size = 0.6) + # optional: show point cloud
  scale_y_log10(
    labels = label_number(accuracy = 0.1),
    breaks = c(0.1, 0.5, 1, 2, 5, 10, 30, 60, 120, 300, 1000)
  ) +
  labs(
    title = "Ride Duration Comparison (log scale)",
    subtitle = "Y-axis in minutes (log10). Outliers compressed to show central tendency.",
    x = "Rider Type",
    y = "Ride Length (minutes, log scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

**Notes:**  
- `scale_y_log10()` compresses the heavy tail so medians and IQRs are more visible.  
- I hid extreme outliers in the boxplot (`outlier.shape = NA`)( but added a faint jitter  cloud in comment so you can still see distribution density if you remove the #).  
- `breaks` include small and large minute values so axis ticks are meaningful.


💭 *Insight:*
Casuals display a far wider duration spread — the classic *tourist vs commuter* contrast.
Members ride frequently but briefly; casuals ride fewer times but linger longer.

---

## Round Trips vs One-Way

```{r roundtrips}
divvy_data %>%
  mutate(is_round_trip = start_station_id == end_station_id) %>%
  group_by(member_casual, is_round_trip) %>%
  summarise(total_rides = n())
```

💭 *Takeaway:*
Casuals record more round trips — likely sightseeing or leisure loops.
Members’ rides are overwhelmingly one-way, consistent with commuting.

---

## Temporal Insights

### Day + Hour Trends

```{r dayhour}
divvy_data %>%
  group_by(member_casual, day_of_week, hour) %>%
  summarise(total_rides = n()) %>%
  ggplot(aes(x = hour, y = total_rides, color = member_casual)) +
  geom_line() +
  facet_wrap(~ day_of_week) +
  labs(title = "Riding Patterns by Day and Hour",
       x = "Hour", y = "Ride Count")
```

💭 *Interpretation:*
Members peak at **8–9 AM** and **5–6 PM** — commute hours.
Casuals peak mid-day and on weekends — pure leisure timing.
Two rhythms emerge: *structured weekdays* vs *spontaneous weekends.*

---

### Seasonal Patterns

```{r season}
divvy_data %>%
  group_by(member_casual, season) %>%
  summarise(total_rides = n(),
            avg_ride_length = mean(ride_length)) %>%
  ggplot(aes(x = season, y = total_rides, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Seasonal Usage by Rider Type",
       x = "Season", y = "Total Rides")
```

💭 *Observation:*
Casual ridership soars in **spring & summer**, drops in winter — weather-sensitive.
Members stay consistent year-round, reflecting reliability and habit.

---

### Weekly Trends

```{r week}
divvy_data %>%
  group_by(member_casual, day_of_week) %>%
  summarise(total_rides = n()) %>%
  ggplot(aes(x = day_of_week, y = total_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Ride Count by Day of Week",
       x = "Day of Week", y = "Number of Rides") +
  theme_minimal()
```

💭 *Observation:*
Members dominate weekdays; casuals surge on weekends — reinforcing work-vs-leisure behavior.

---

## Spatial Insights

### Top 10 Start Stations

```{r topstations}
top_stations <- divvy_data %>%
  group_by(member_casual, start_station_name) %>%
  summarise(rides = n()) %>%
  arrange(member_casual, desc(rides)) %>%
  slice_head(n = 10)

ggplot(top_stations,
       aes(x = reorder(start_station_name, rides),
           y = rides, fill = member_casual)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~ member_casual, scales = "free_y") +
  labs(title = "Top 10 Start Stations per Rider Type",
       x = "Station", y = "Rides")
```

💭 *Insight:*
Members cluster around **business districts and transit hubs**.
Casuals favor **tourist zones and waterfronts** — ideal spots for day-pass marketing.

---

# STRATEGIC TAKEAWAYS

| Insight Area      | Observation                                              | Business Implication                                  |
| ----------------- | -------------------------------------------------------- | ----------------------------------------------------- |
| **Ride Duration** | Casual rides are ≈ 3× longer                             | Promote annual memberships for frequent leisure users |
| **Peak Times**    | Members commute mornings/evenings; casuals ride weekends | Offer weekday discounts to casuals                    |
| **Seasonality**   | Casual rides surge in summer                             | Launch pre-summer marketing campaigns                 |
| **Top Stations**  | Casuals use tourist-heavy stations                       | Place ads / QR banners at these hubs                  |
| **Round Trips**   | Casuals take more round trips                            | Introduce “unlimited short-ride” plans                |

💭 *Final Thoughts:*
**Members = commuters; Casuals = tourists and leisure seekers.**
Push membership incentives during warmer months when casual ridership peaks — ideally before summer weekends.
Highlight **cost savings, convenience, and priority perks** to nudge casual riders into annual membership tiers.



---



# **Recommendations**

---


## **1️⃣ Seasonal Membership Campaign (Late Spring – Early Summer)**

**What:**
Launch a targeted membership promotion focused on converting casual riders during peak leisure months.

**How:**

* Advertise membership perks tied to leisure use (ride discounts, scenic-station offers, partner attraction perks).
* Use QR codes at high-traffic scenic stations for instant sign-up.

**When:**

* Begin advertising in **late April**
* Run campaign through **July**, when casual ridership is highest.

**By Whom:**

* **Marketing Team** (campaign design, messaging, creatives)
* **Partnerships Team** (tourist sites, attractions, scenic zones)
* **Operations Team** (QR placement at stations)

---

## **2️⃣ Public Transport Advertising for Casual Riders**

**What:**
Promote Cyclistic membership on public transport channels frequently used by casual riders.

**How:**

* Place ads with QR codes inside buses, trains, and metro stations connected to tourist and weekend-heavy areas.
* Highlight convenience and cost savings for multi-modal travelers.

**When:**

* Run continuously, with increased frequency on **weekends** and **tourist season** (May–September).

**By Whom:**

* **Marketing Team** (ad design, CTA optimisation)
* **Transit Advertising Partners** (bus, metro, rail networks)
* **Brand Outreach Team** (negotiating placements)

---

## **3️⃣ Add Extra Ride-Time Benefits to Annual Membership**

**What:**
Provide additional free ride-time hours as a built-in benefit for annual members.

**How:**

* Offer monthly bonus minutes or extended ride duration before overage fees.
* Promote these perks during onboarding and all seasonal campaigns.

**When:**

* Introduce the benefit at the start of **summer** (June)
* Maintain year-round as a permanent membership feature.

**By Whom:**

* **Product Team** (membership feature design)
* **Finance Team** (cost modeling)
* **Marketing Team** (promotion and messaging)

---

If you'd like, I can also format this as a **one-page executive summary**, a **slide deck**, or insert it cleanly into your final PDF/HTML report.

# CONCLUSION

This analysis explored how Cyclistic’s members and casual riders differ in their riding behavior across time, season, and location.  
Members ride frequently, briefly, and predictably — reflecting daily commuters who depend on Cyclistic for practicality and routine.  
Casual riders, on the other hand, ride longer, mostly on weekends and during warmer months, often looping back to where they started — echoing leisure, tourism, and exploration.  

These patterns reveal two distinct motivations: **efficiency vs experience**.  
Bridging that gap through well-timed seasonal promotions, flexible membership plans, and on-site marketing at tourist stations can transform occasional riders into loyal members.  

